Forecasting the 2024 U.S. Presidential Election

### 2024-11-02 paper/paper.qmd
model refinement with discussion refinement

@@

To forecast the winner of the 2024 U.S. Presidential election between Kamala Harris and Donald Trump, we performed the following series of steps, focusing on the distribution of electoral votes across states based on predicted polling percentages. This approach allows for an estimation of each candidate's electoral college support under the U.S. voting system.

1. Predicting State-Level Polling Percentages: We used final multiple linear regression models for both candidates, Harris and Trump. These models generated predicted polling percentages for each candidate across different states, capturing their expected levels of support based on the poll data used.

2. Averaging Predicted Support by State: For both Harris and Trump, we calculated the average predicted polling percentage within each state. This step provides a state-level summary of each candidate's support based on our predictions, simplifying comparisons between them in each state.

3. Assigning Electoral Votes: We then merged our state-level predictions with the distribution of electoral votes, as assigned by state in the Electoral College. This dataset provided the electoral vote count for each state, including special allocations for districts in Maine and Nebraska, which can split their electoral votes by congressional district.

4. Determining State Winners: Using the averaged predictions, we identified the winner in each state by comparing the predicted support percentages. Harris or Trump was deemed the winner in a state if their predicted polling percentage exceeded that of their opponent.

5. Aggregating Electoral Votes: Finally, we summed the electoral votes for each candidate across all states where they were predicted to win. This yielded total electoral vote counts for both Harris and Trump, which are essential for determining the predicted winner under the U.S. system, where 270 electoral votes are required to secure the presidency.

6. Interpreting the Results: With Harris predicted to receive 216 electoral votes compared to Trump’s 147, Harris demonstrates a stronger position in the Electoral College, though neither candidate reaches the 270-vote threshold required for a decisive victory. This outcome suggests that, while a clear electoral win is not forecasted, Harris holds a better chance of winning the election based on her lead in projected electoral votes. This lead places her closer to the threshold and may indicate a competitive advantage should undecided states or close races lean in her favor.

This electoral vote aggregation based on state-level predictions provides an interpretation of how polling data (@fig-predicted-vote-pct), analyzed through multiple linear regression models, translates into likely electoral support for each candidate. It effectively models potential election outcomes while respecting the structure of the Electoral College​​.

@@

## Key Findings

This study set out to forecast the outcome of the 2024 U.S. Presidential Election between Kamala Harris and Donald Trump by developing a multiple linear regression (MLR) model based on comprehensive polling data. Importantly, the same MLR model was applied to both candidates, ensuring consistency in how predictors influenced the predicted support percentages for each.

The MLR model incorporated several significant predictors of voter support, including state-level factors, pollster ratings (pollscore), transparency scores, and sample size. By using the same model for both Harris and Trump, we were able to directly compare how these variables impacted each candidate's predicted support across different states.

For both Kamala Harris and Donald Trump, the model highlighted the substantial impact of state-level variations on their predicted support. Certain states showed higher predicted support for Harris, such as California and Maryland, which historically lean Democratic. Conversely, states like Indiana and Missouri exhibited higher predicted support for Trump, aligning with their Republican-leaning tendencies. The consistency of the model across both candidates allowed for a clear comparison of how state demographics and historical voting patterns influenced voter preferences.

Pollster reliability indicators, such as numeric grades and transparency scores, were significant predictors for both candidates. This suggests that polls conducted by more credible organizations tended to report more accurate and potentially higher support levels for each candidate. The sample size also played a role, although its impact varied slightly between the two candidates. By using the same model, we ensured that these predictors were weighted equally in the analysis of both Harris's and Trump's support levels.

When aggregating the state-level predictions to simulate the Electoral College outcomes, the findings suggest a competitive race. Harris is projected to receive 216 electoral votes, while Trump is predicted to secure 147 electoral votes. Neither candidate reaches the 270-vote threshold required to win the presidency, emphasizing the potential for a closely contested election. The remaining electoral votes are likely concentrated in battleground states, where voter preferences are more volatile and could ultimately determine the election outcome.

## Model Strengths

A primary strength of this analysis lies in the consistent application of the same MLR model to both candidates. This approach ensures that the comparison between Harris and Trump is based on the same criteria and that any differences in predicted support are attributable to variations in the data rather than differences in the modeling approach.

By incorporating multiple predictors—including state-level data, pollster ratings, transparency scores, and sample size—the model provides a nuanced understanding of the factors influencing voter preferences. This multifaceted approach allows for more accurate predictions compared to models relying solely on national-level polling data or a single set of predictors.

The use of stepwise model selection and checks for multicollinearity enhanced the robustness of the MLR model. Stepwise selection optimized the predictor variables, ensuring that only those contributing significantly to the model were included. Checking for multicollinearity through the Variance Inflation Factor (VIF) helped prevent unreliable estimates due to correlated predictors, thereby improving the model's validity for both candidates.

Moreover, translating the predicted support percentages into Electoral College projections adds practical value to the analysis. Since the U.S. Presidential Election is determined by electoral votes rather than the popular vote, this approach offers a more realistic forecast of the election outcome. Using the same model for both candidates ensures that the Electoral College projections are directly comparable.

@@

Despite the strengths, several limitations need to be acknowledged:

- Reliance on Polling Data Quality: The accuracy of the model is inherently dependent on the quality of the polling data used. While the analysis accounted for pollster ratings and transparency scores, inherent biases and methodological differences between polls could still affect the results. Sampling errors, non-response biases, and underrepresentation of certain demographic groups are persistent issues in polling data that can skew predictions for both candidates.

- Static Snapshot of Voter Preferences: The model provides a static view based on the most recent polling data available. It does not account for the dynamic nature of election campaigns, where voter preferences can shift rapidly due to various factors such as political events, debates, or emerging issues. This limitation means the model may not accurately predict future changes in voter sentiment leading up to the election for either candidate.

- Exclusion of Other Influential Factors: The analysis did not incorporate other variables that can significantly impact election outcomes, such as economic indicators (e.g., unemployment rates, GDP growth), social movements, campaign strategies, or voter turnout efforts. The absence of these factors may limit the model's ability to fully capture the complexities of voter behavior for both Harris and Trump.

- Electoral College Simplifications: The Electoral College simulation assumes that the candidate with the higher predicted support in a state wins all its electoral votes (except for Maine and Nebraska). This winner-takes-all approach does not account for the proportional allocation of electoral votes in those two states or potential variations in voter turnout that could affect the actual results.

## Future Research 
To enhance the predictive power and accuracy of election forecasting models, future research could consider the following approaches:

- Incorporate Time-Series Analysis: Introducing a time-series component would allow the model to account for trends and shifts in voter preferences over time. This dynamic approach could capture the impact of events such as debates, policy announcements, or external factors on voter sentiment for both candidates.

- Expand Predictive Variables: Including additional variables such as economic indicators, demographic data, and social media sentiment analysis could provide a more comprehensive view of the factors influencing voter behavior. Integrating data on unemployment rates or consumer confidence might reveal economic influences on candidate support.

- Voter Turnout Models: Developing models to predict voter turnout could significantly improve election forecasts. Turnout can vary widely between elections and is often influenced by voter enthusiasm, mobilization efforts, and barriers to voting. Incorporating turnout probabilities could adjust the predicted support percentages to reflect more realistic electoral scenarios.

- Advanced Modeling Techniques: Exploring advanced statistical methods or machine learning algorithms might capture complex nonlinear relationships between predictors and voter support. Techniques such as random forests, gradient boosting, or neural networks could uncover patterns not detectable through linear regression.

- Account for Electoral Nuances: With variations in how electoral votes are allocated and the potential impact of third-party candidates or ranked-choice voting in some states, future models should consider these electoral nuances. Simulating different voting systems could provide insights into how alternative methods might affect election outcomes.

- Cross-Validation with Alternative Data Sources: Validating the model against alternative data sources, such as exit polls or historical voting patterns, could test its robustness. Cross-validation helps ensure that the model is not overfitted to the polling data used and can generalize to different datasets.

@@

```{r, out.width="90%", out.height="70%", fig.align="center"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-predicted-vote-pct
#| fig-cap: Predicted Vote Percentages by State
# knitr::include_graphics(here::here(("improved_vote_percentages_by_state.png")))

>>>>

```{r}
#| echo: false
#| message: false
#| warning: false
# Load necessary libraries
library(reshape2)
library(ggplot2)

# Load final models for Harris and Trump
mlr_harris_model_final <- readRDS(here::here("models/mlr_harris_model_final.rds"))
mlr_trump_model_final <- readRDS(here::here("models/mlr_trump_model_final.rds"))

# Predict poll percentages for Kamala Harris and Donald Trump
harris_data$predicted_pct_harris <- predict(mlr_harris_model_final, newdata = harris_data)
trump_data$predicted_pct_trump <- predict(mlr_trump_model_final, newdata = trump_data)

```{r}
#| echo: false
#| message: false
#| warning: false
# Define electoral votes for each state (and Washington, D.C.)
# Correct the electoral_votes data frame
electoral_votes <- data.frame(
  state = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", 
    "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
    "Kansas", "Kentucky", "Louisiana", 
    # Maine entries
    "Maine At-Large", "Maine CD-1", "Maine CD-2", 
    "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", 
    # Nebraska entries
    "Nebraska At-Large", "Nebraska CD-1", "Nebraska CD-2", "Nebraska CD-3", 
    "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", 
    "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", 
    "Wyoming"
  ),
  electoral_votes = c(
    9, 3, 11, 6, 55, 9, 7, 3, 3, 29, 16, 4, 4, 20, 11, 6, 6, 8, 8,
    # Maine electoral votes
    2, 1, 1,
    10, 11, 16, 10, 6, 10, 3,
    # Nebraska electoral votes
    2, 1, 1, 1,
    6, 4, 14, 5, 29, 15, 3,
    18, 7, 7, 20, 4, 9, 3,
    11, 38, 6, 3, 13, 12, 5, 10,
    3
  )
)

# Aggregate predictions by state for Harris and Trump
harris_state_avg <- aggregate(predicted_pct_harris ~ state, data = harris_data, FUN = mean)
trump_state_avg <- aggregate(predicted_pct_trump ~ state, data = trump_data, FUN = mean)

# Merge predictions for both candidates
prediction_comparison <- merge(harris_state_avg, trump_state_avg, by = "state", all.x = TRUE)

```
```{r}
#| echo: false
#| message: false
#| warning: false
# Identify any states missing from the prediction data
missing_states <- setdiff(electoral_votes$state, prediction_comparison$state)

# Assign 50% support to both candidates for missing states
missing_states_data <- data.frame(
  state = missing_states,
  predicted_pct_harris = 50,
  predicted_pct_trump = 50
)

# Append missing states data to prediction_comparison
prediction_comparison <- rbind(prediction_comparison, missing_states_data)

# Merge electoral votes into the prediction_comparison to ensure each state has electoral votes
prediction_comparison <- merge(prediction_comparison, electoral_votes, by = "state", all.x = TRUE)

```
```{r}
#| echo: false
#| message: false
#| warning: false
# Determine the winner for each state based on predicted percentages
set.seed(304)
prediction_comparison$winner <- ifelse(
  prediction_comparison$predicted_pct_harris > prediction_comparison$predicted_pct_trump, 
  "Harris", 
  ifelse(prediction_comparison$predicted_pct_trump > prediction_comparison$predicted_pct_harris, 
         "Trump", 
         sample(c("Harris", "Trump"), nrow(prediction_comparison), replace = TRUE))
)

# Calculate total electoral votes for each candidate
harris_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Harris"], na.rm = TRUE)
trump_electoral_votes <- sum(prediction_comparison$electoral_votes[prediction_comparison$winner == "Trump"], na.rm = TRUE)

# Create a table with electoral votes
electoral_votes_table <- data.frame(
  Candidate = c("Harris", "Trump"),
  Electoral_Votes = c(harris_electoral_votes, trump_electoral_votes)
)

# Display the table with kable
kable(electoral_votes_table, caption = "Electoral Votes for Each Candidate", col.names = c("Candidate", "Electoral Votes"))
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-predicted-vote-pct
#| fig.height: 5
#| fig.cap: "Predicted Vote Percentages by State"

# Load necessary libraries
library(ggplot2)
library(reshape2)
library(dplyr)
library(patchwork)

# Melt data for easier plotting
prediction_melted <- melt(
  prediction_comparison, 
  id.vars = c("state", "electoral_votes", "winner"),
  measure.vars = c("predicted_pct_harris", "predicted_pct_trump")
)

# Rename melted variables for readability
colnames(prediction_melted)[colnames(prediction_melted) == "variable"] <- "candidate"
prediction_melted$candidate <- ifelse(
  prediction_melted$candidate == "predicted_pct_harris", 
  "Harris", 
  "Trump"
)

# Calculate the midpoint to split the states
n_states <- length(unique(prediction_melted$state))
midpoint <- ceiling(n_states / 2)

# Order states by electoral votes in descending order for better visualization
states_order <- prediction_melted %>%
  distinct(state, electoral_votes) %>%
  arrange(desc(electoral_votes)) %>%
  pull(state)

# Split the states into two groups
left_states <- states_order[1:midpoint]
right_states <- states_order[(midpoint + 1):length(states_order)]

# Create two separate datasets
prediction_left <- prediction_melted %>%
  filter(state %in% left_states) %>%
  mutate(state = factor(state, levels = left_states))  # Ensure consistent ordering

prediction_right <- prediction_melted %>%
  filter(state %in% right_states) %>%
  mutate(state = factor(state, levels = right_states))  # Ensure consistent ordering

# Function to create individual plots to avoid repetition
create_plot <- function(data, show_legend = FALSE) {
  ggplot(data, aes(x = reorder(state, electoral_votes), y = value, fill = candidate)) +
    geom_bar(
      stat = "identity", 
      position = position_dodge(width = 0.9), 
      width = 0.5  # Reduced bar width for more spacing
    ) +
    coord_flip() +
    labs(
      x = NULL,
      y = "Predicted Vote Percentage (%)"
    ) +
    scale_fill_manual(values = c("Harris" = "#87CEFA", "Trump" = "#F08080")) +
    theme_minimal() +
    theme(
      text = element_text(size = 9),  # Reduced overall text size
      axis.text.y = element_text(size = 8, margin = margin(r = 5)),  # Smaller state labels
      axis.text.x = element_text(size = 9),
      legend.title = element_blank(),
      legend.position = ifelse(show_legend, "top", "none"),  # Place legend only for the first plot
      legend.text = element_text(size = 9),
      panel.grid.major.y = element_blank(),
      plot.margin = margin(l = 10, r = 15, t = 10, b = 10)  # Adjusted margins
    ) +
    geom_text(
      aes(label = round(value, 1)),
      position = position_dodge(width = 0.9),
      hjust = -0.2,    # Adjusted horizontal position for labels
      size = 1.5,      # Smaller label size
      show.legend = FALSE
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.2)),  # Increased expansion for labels
      limits = c(0, 75)  # Adjust based on your data
    )
}

# Create the left plot with legend
plot_left <- create_plot(prediction_left, show_legend = TRUE)

# Create the right plot without legend
plot_right <- create_plot(prediction_right, show_legend = FALSE)

# Combine the plots side by side with legend on top
combined_plot <- plot_left + plot_right +
  plot_layout(guides = "collect") &  # Collects the legend at the top
  theme(legend.position = "top")  # Places the combined legend at the top

# Print the combined plot
combined_plot

To project the outcome of the 2024 U.S. Presidential election between Kamala Harris and Donald Trump, we conducted a statistical analysis that integrates predicted state-level polling percentages with the allocation of electoral votes. This methodology employs multiple linear regression models to estimate each candidate's support and translates these predictions into potential Electoral College results, adhering to the structure of the U.S. electoral system.

1. **Prediction of State-Level Polling Percentages**: We utilized finalized multiple linear regression models for both Kamala Harris and Donald Trump. These models were applied to their respective datasets to generate predicted polling percentages for each candidate across various states. The predictions reflect the expected levels of support based on historical polling data and relevant covariates included in the models.

2. **Aggregation of Predicted Support by State**: For each candidate, we computed the average predicted polling percentage within each state. This aggregation provides a concise state-level summary of anticipated support, facilitating direct comparisons between the candidates in each state (see Figure @fig-predicted-vote-pct).

3. **Integration of Electoral Vote Allocations**: We established a dataset detailing the electoral vote distribution for each state, including the distinct allocations for Maine and Nebraska, which can split their electoral votes by congressional district. This dataset was merged with the state-level predicted support data to align electoral votes with the corresponding predicted support.

4. **Addressing Missing Data**: We identified states lacking predicted polling data and assigned a neutral predicted support of 50% to both candidates in these states. This assumption accounts for the absence of data while ensuring that all states are included in the analysis.

5. **Determination of State Winners**: Based on the aggregated predicted percentages, we determined the projected winner in each state. A candidate was designated as the winner if their predicted polling percentage exceeded that of their opponent. In cases where the predicted percentages were equal, we randomly assigned the winner using a stochastic approach to reflect the uncertainty inherent in tied predictions.

6. **Calculation of Total Electoral Votes**: We summed the electoral votes from all states won by each candidate to compute their total electoral vote counts. This calculation is critical, as securing at least 270 electoral votes is necessary to win the presidency under the U.S. electoral system.

7. **Interpretation of Results**: The analysis indicated that Kamala Harris is projected to receive a total of **306 electoral votes**, while Donald Trump is projected to receive **232 electoral votes**. Harris surpasses the 270-vote threshold required for victory, suggesting a strong position in the Electoral College based on our predictions. This outcome implies that Harris is likely to win the election, given her substantial lead in projected electoral votes.

This statistical approach effectively bridges the gap between state-level polling data and the Electoral College outcome. By employing multiple linear regression models to predict polling percentages and integrating these predictions with the electoral vote framework, we provide a nuanced analysis of potential election results. This methodology respects the intricacies of the U.S. electoral system and offers valuable insights into how predicted voter support may translate into electoral success.

@@

## Summary of Findings

In this paper, we developed statistical models to forecast the outcome of the 2024 United States presidential election between Kamala Harris and Donald Trump. Utilizing polling data from FiveThirtyEight's 2024 Presidential Election Forecast Database, we analyzed factors influencing candidate support through linear and multiple linear regression models. We incorporated variables such as pollster rating, transparency score, sample size, and state-level data to predict the percentage of support for each candidate across different states. By aggregating these state-level predictions and integrating electoral vote allocations, we projected that Kamala Harris would receive 306 electoral votes, surpassing the 270 required to win the presidency.

## Insights on Polling Methodologies and Voter Support

Our analysis indicates that pollster rating and state-level factors significantly influence reported support for both candidates. For Kamala Harris, the multiple linear regression model shows that higher pollster ratings and certain states are associated with increased support percentages. This suggests that polls conducted by more reliable pollsters may report higher support for Harris, possibly due to methodological rigor or sampling techniques that capture her voter base more accurately. For Donald Trump, pollster transparency and state differences play a notable role, indicating that how openly pollsters disclose their methodologies can affect the reported support for Trump.

## Implications for Electoral Forecasting

The projection of Kamala Harris receiving 306 electoral votes underscores the importance of incorporating state-level analyses in electoral forecasting. Our findings highlight that national polling averages may not sufficiently capture the complexities of the Electoral College system. By focusing on state-specific predictions and accounting for variables that influence voter support in different regions, forecasters can achieve more accurate predictions. This approach acknowledges the heterogeneity of the electorate and the pivotal role of battleground states in determining election outcomes.

@@

Despite the strengths of our modeling approach, several limitations must be acknowledged. First, the reliance on historical polling data and pollster ratings assumes that past performance is indicative of future accuracy, which may not hold true if polling methodologies or voter behaviors change significantly. Second, the underrepresentation of certain states due to sparse polling data may affect the reliability of our state-level predictions. Assigning a default 50% support in states with missing data introduces uncertainty and may not reflect actual voter preferences. Additionally, the models may not fully account for dynamic factors such as late-breaking events, shifts in voter sentiment, or turnout variations that can influence election results.

## Future Research and Recommendations

To enhance the accuracy of electoral forecasts, future research should consider incorporating real-time data and alternative data sources such as social media trends, economic indicators, and demographic shifts. Employing more sophisticated modeling techniques, such as hierarchical Bayesian models or machine learning algorithms, may capture complex interactions and non-linear relationships among variables. Furthermore, improving polling methodologies to address issues of response bias and sample representativeness is essential. Collaborations between pollsters, statisticians, and political scientists can foster the development of more robust predictive models that adapt to the evolving electoral landscape.

@@
